Project Description -- Erosion, Transport, Anomalous Diffusion, and Collapse in Granular and Porous Media

### 2019-11-14 docs/proposals/NSF/ProjDesc/ProjDesc.tex
picking away at proposal

\item Additionally effects, such as sedimentation, transport, and compaction constantly work to reconfigure the medium as its constituents and boundaries degrade.
\end{itemize}
These challenges require innovative computational techniques, some of which have been developed recently~\cite{Quaife2018} and others that await resolution. An ultimate goal of this proposal is to achieve high-fidelity simulations of $\mathcal{O}(100-1000)$ mobile bodies undergoing simultaneous erosion, transport, sedimentation, and compaction, with the computational efficiency required to run large ensembles for statistical analysis. Dynamic erosion of porous media is a highly active area of research, as closely related laboratory experiments on erosion in colloidal systems are already being explored in the Datta Lab at Princeton~\cite{bizmark2019multiscale}. These experiments a great opportunity for comparison against our numerical simulations. Furthermore, we propose performing a set of experiments conducted at the nearby the Geophysical Fluid Dynamics Institute, which would help to guide, calibrate, and validate the computations.

\todo[inline]{Bryan: Seems like a natural place to add some letters of
collaboration. Kevin S and/or Ming might be 2. Not sure if you have
someone at Princeton that you would ask for one from?}
\todo[inline]{Nick: Good point. How do you include collaboration letters in the text? Do you cite them like a reference? I will ask Kevin or Ming for a collaboration letter. I don't think we need to ask the people at Princeton too; I meant that to mainly indicate it is an exciting, active area of research and so I rephrased the sentence.}

@@

We first summarize governing equations underlying the proposed work.
At the finest scale, the fluid flow intervening between individual grains of a porous medium is governed by the incompressible Stokes equations
\begin{equation}
\label{StokesEq}
\begin{split}
\mu \Delta \uu = \grad p,	&\hspace{20pt} \xx \in \Omega, \\
\grad \cdot \uu = 0, 		&\hspace{20pt} \xx \in \Omega,
\end{split}
\end{equation}
where $\uu$ is the velocity field, $p$ the pressure field, and $\mu$ the fluid viscosity. Linear elliptic PDEs, such as the Stokes equations, can be reformulated as a boundary integral equation (BIE).  BIEs have several numerical advantages including a dimension reduction, automatic mass conservation, and high-order accuracy. An equivalent BIE formulation of the Stokes equations in Eq.~\eqref{StokesEq} is
\begin{align}
  \label{eqn:BIE}
  \ff(\xx) = -\frac{1}{2}\eeta(\xx) + \frac{1}{\pi} \int_{\bd\Omega}
    \frac{\rr \cdot \nn}{\rho^2} \frac{\rr \otimes \rr}{\rho^2}
    \eeta(\yy) ds_\yy, \quad \xx \in \bd\Omega
\end{align}
where $\ff$ is a Dirichlet boundary condition, $\rr = \xx - \yy$, $\rho = \|\rr\|$, and $\nn$ is the outward unit normal of $\bd\Omega$.  Once~\eqref{eqn:BIE} is solved for the density function $\eeta$, important quantities such as the velocity, pressure, vorticity, and deformation tensor can be computed in the fluid bulk.  For example, the velocity at $\xx \in \Omega$ is 
\begin{align*}
  \uu(\xx) = \frac{1}{\pi} \int_{\bd\Omega}
    \frac{\rr \cdot \nn}{\rho^2} \frac{\rr \otimes \rr}{\rho^2}
    \eeta(\yy) ds_\yy.
\end{align*}

@@

 Figure \ref{fig1} shows preliminary results of multiple bodies undergoing mechanical erosion in Stokes flow. The figure shows 80 bodies of initially random size and position, undergoing erosion. This figure extends previously reported 50-body simulations to 80 bodies~\cite{Quaife2018} due to a recently developed barycentric interpolation scheme~\cite{bar2014, bar-wu-vee2015}, which enables closer contact between bodies and thus more bodies.
The color shows the (normalized) vorticity field surrounding the bodies. Vorticity, since it reduces to shear on solid interfaces, provides a convenient way to visualize local erosion rates, as well as the local flow intensity. Observe that erosion not only reduces the size of solid bodies, but also alters their shapes substantially. The bodies tend to become somewhat polygonal: corners develop connected by relatively flat faces. The number of faces does not appear to be easily predicted, but rather depends on the interaction with neighboring bodies as mediated by the Stokes flow. Notice that, as erosion proceeds, relatively straight channels tend to develop between the bodies. Certain channels that are initially larger can transmit more flow, which promotes local erosion rates and further widens these selected channels. In this way, erosion creates a runaway process in which small differences in initial channel size become amplified. The channelization process is made more evident by visualization of the velocity magnitude, as in Fig.~\ref{fig2}.

@@

An immediate goal of the proposed work is extend the range of physical effects captured in these simulations. A particular challenge of a BIE formulation is developing quadrature for nearly-singular integrands that arise when bodies are close. To address this problem, we have very recently extended a Barycentric quadrature~\cite{abr2014, bar-wu-vee2015} and have successfully simulated erosion with grains that are initially separated by 1\% of an arclength spacing. This quadrature approach has the advantage that it is a non-intrusive modification of the trapezoid rule, allowing us to achieve an error bound that independent of the grain configuration. Near-term plans are to include flow transport and gravitational sedimentation of the eroding grains. For this extension, contact forces between bodies must be included to eliminate unphysical contact that is inevitable due to discretization errors.

@@

\todo[inline]{Bryan, could you add some details and references to the paragraph below?}

Next, we plan to extend the simulation framework to handle chemical dissolution of solid bodies via physical laws \eqref{Ceq}--\eqref{DissVn}. For this goal, much of the computational infrastructure, such as the boundary-integral solver for the Stokes equations, is already in place, but the coupling to the advection-diffusion equations~\eqref{Ceq} and interface law~\eqref{DissVn} would need to be implemented. Once these pieces are in place, simulations of simultaneous erosion and dissolution, while also including transport and sedimentation, would finally become possible. Such a framework could accurately simulate the geophysically relevant scenario of karst conduits with embedded granular media. The gradual changes in the medium and karst boundaries due to erosion and dissolution could create conditions favorable for mechanical failure or collapse of the structure, for example sinkholes.
\todo[inline]{Some of this stuff is already in the Transport section
such as the Laplace transform idea}

\todo[inline]{Nick to Bryan: These changes you made look good to me. Did you want to add anything else, or is it okay to delete the above todo notes?}

@@

\begin{wrapfigure}[]{r}{0.3\textwidth}
  \includegraphics[width=0.3\textwidth]{figs/100b_t100tracer}
  \caption{\label{fig:100tracers} A visualization of the streamlines
  with 200 tracers passing through an eroded geometry with a porosity of
  0.63. The tracers are initialized at the left end of the channel.}
\end{wrapfigure}

Transport through eroded media is a direct consequence of the pore
structure. Qualitative properties of an eroded media, such as channels
of high porosity can be modest in volume fraction, but still be
responsible for a large portion of the flux~\cite{Quaife2018}. 
\todo[inline]{There is no verb in the above sentence. Is there a word left out?}
To increase understanding of flow through eroded media, we will simulate
and statistically analyze transport through eroded geometries.

In an eroded geometry $\Omega$, the dimensionless
concentration $c$ is governed by
\begin{align}
  \pd{c}{t} + \uu \cdot \nabla c = \frac{1}{\Pe} \Delta c, 
  \label{eqn:advectionDiffusion}
\end{align}
where $\Pe$ is the dimensionless Peclet number. Here we assume that the
concentration $c$ is affected by the eroded geometry, but the geometry
is not altered further by the concentration. In the infinite Peclet
number limit, the concentration is governed by pure advection and the
characteristics of the concentration field are identical to the
streamlines (Figure~\ref{fig:100tracers}).
\begin{wrapfigure}[]{l}{0.4\textwidth}
  \includegraphics[width=0.4\textwidth]{figs/tort_eulerian100}
  \caption{\label{fig:100tortuosity} The tortuosity of an eroding
  geometry. The tortuosity can be computed using two different formulas
  (red vs.~blue marks). The red square is the tortuosity in
  Figure~\ref{fig:100tracers}. Their is qualitative agreement between
  our numerical simulations and a power-law model (black dashed
  line)~\cite{mat-kha-koz2008}.}
\end{wrapfigure}

The $j^{th}$ trajectories, or characteristics, is governed by
$\dot{\ss}_j(t) = \uu(\ss_j(t))$. Once these trajectories are computed,
using for example a high-order Runge-Kutta method, a statistical
analysis can be used to describe the transport. Two of the most important
measurements are the tortuosity and the anomalous dispersion, both which
are defined in terms of the length of the trajectories
\begin{align}
  \lambda_j(t) = \int_{0}^{t} \|\ss'(t')\| dt'.
\end{align}
Once $N_p$ trajectories are computed, the tortuosity is the
dimensionless number
\begin{align}
  T = \frac{1}{d} \left(\int_{S} u_1(x_0,y_0)\lambda(y_0) dy_0\, \right)
  \Bigg/ \left(\int_S u_1(x_0,y_0)\, dy_0 \right),
\end{align}
where $S$ is the vertical cross-section at $x=x_0$, $d$ is its length,
$u_1$ is the horizontal velocity, and $\lambda(y_0)$ is the length of
the trajectory starting at $(x_0,y_0)$ and ending at a vertical
cross-section at the other end of the channel. Relationships between the
tortuosity and porosity of a geometry have been developed by other
groups~\cite{kop-kat-tim1996, dud-koz-mat2011, mat-kha-koz2008}, and
preliminary results indicate that such models might be appropriate for
geometries whose porosity is increasing under the process of erosion
(Figure~\ref{fig:100tortuosity}).

While the tortuosity characterizes the amount of winding of streamlines,
the dispersion characterizes the spreading of the trajectories.
Dispersion can be descried using the first- and second-ensemble moments
of the trajectories
\begin{align}
  \langle \lambda \rangle (t) = \frac{1}{N_p} \sum_{j=1}^{N_p}
    \lambda_j(t), \quad \sigma_\lambda^{2}(t) = \frac{1}{N_p}
    \sum_{j=1}^{N_p} (\lambda_j(t) - \langle \lambda \rangle(t))^2
\end{align}
Then, the standard deviation $\sigma_\lambda$ characterizes the
dispersion, and long-time behavior typically results in a power-law
scaling $\sigma_\lambda \sim t^{\alpha}$. The power $\alpha$, which
quantifies the long-time spreading, is $1$ in a ballistic regime which
occurs at early times when the trajectories have not explored much of
the porous geometry. However, once the trajectories have traversed a few
grains, dispersion in porous media is often super-dispersive with
$\alpha \in (0.5,1)$~\cite{}. In Figure~\ref{fig:100dispersion}, we
investigate spreading at several different porosities of a porous
geometry that is initialized with 100 grains.
\begin{wrapfigure}[]{r}{0.4\textwidth}
  \includegraphics[width=0.4\textwidth]{figs/100b_second_moment_long_ref}
  \caption{\label{fig:100dispersion} The particle spreading,
  characterized with the standard deviation $\sigma_\lambda(t)$ at 6
  different porosities. At early times, the line of slope 1 (black
  curve) indicates a ballistic motion. Then, at later times, the
  dispersion becomes super-dispersive at a rate indicated by the slope
  of the black dashed-dotted line.}
\end{wrapfigure}

@@

	In very preliminary work, PI Moore has been running simplified simulations with a current graduate student. These preliminary simulations are based on a hybrid continuum-discrete model for the permeability field and grain particle field respectively. They utilize the homogenized description of the flow field in \eqref{DarcyEq} and incorporate the physical effects of:
\begin{itemize}
\item Transport of grains due to seepage force from the surrounding fluid flow.
\item Gravitational sedimentation of grains.
\item Feedback of grain-distribution onto permeability field.
\item Contact and cohesive forces between grains.
\end{itemize}
Preliminary investigation suggests that this last mechanisms of cohesion is essential to model sinkhole collapse in the system.

>>>>

\item Additionally effects, such as sedimentation, transport, and compaction constantly work to reconfigure the medium as its constituents and boundaries degrade.
\end{itemize}
These challenges require innovative computational techniques, some of which have been developed recently~\cite{Quaife2018} and others that await resolution. An ultimate goal of this proposal is to achieve high-fidelity simulations of $\mathcal{O}(100-1000)$ mobile bodies undergoing simultaneous erosion, transport, sedimentation, and compaction, with the computational efficiency required to run large ensembles for statistical analysis. Dynamic erosion of porous media is a highly active area of research, as closely related laboratory experiments on erosion in colloidal systems are already being explored in the Datta Lab at Princeton~\cite{bizmark2019multiscale}. These experiments a great opportunity for comparison against our numerical simulations. Furthermore, we propose performing a set of experiments conducted at the nearby the Geophysical Fluid Dynamics Institute, which would help to guide, calibrate, and validate the computations (see letter of collaboration).

@@

We first summarize governing equations underlying the proposed work.
At the finest scale, the fluid flow intervening between individual grains of a porous medium is governed by the incompressible Stokes equations
\begin{equation}
\label{StokesEq}
\begin{split}
\mu \Delta \uu = \grad p,	&\hspace{20pt} \xx \in \Omega, \\
\grad \cdot \uu = 0, 		&\hspace{20pt} \xx \in \Omega,
\end{split}
\end{equation}
where $\uu$ is the velocity field, $p$ the pressure field, and $\mu$ the fluid viscosity. Linear elliptic PDEs, such as the Stokes equations, can be reformulated as a boundary integral equation (BIE).  BIEs have several numerical advantages including a dimension reduction, automatic mass conservation, and high-order accuracy. An equivalent BIE formulation of the Stokes equations in Eq.~\eqref{StokesEq} is
\begin{align}
  \label{eqn:BIE}
  \ff(\xx) = -\frac{1}{2}\eeta(\xx) + \frac{1}{\pi} \int_{\bd\Omega}
    \frac{\rr \cdot \nn}{\rho^2} \frac{\rr \otimes \rr}{\rho^2}
    \eeta(\yy) ds_\yy, \quad \xx \in \bd\Omega
\end{align}
where $\ff$ is a Dirichlet boundary condition, $\rr = \xx - \yy$, $\rho = \|\rr\|$, and $\nn$ is the outward unit normal of $\bd\Omega$.  Once~\eqref{eqn:BIE} is solved for the density function $\eeta$, important quantities such as the velocity, pressure, vorticity, and deformation tensor can be computed in the fluid bulk.  For example, the velocity at $\xx \in \Omega$ is 
\begin{align}
  \uu(\xx) = \frac{1}{\pi} \int_{\bd\Omega}
    \frac{\rr \cdot \nn}{\rho^2} \frac{\rr \otimes \rr}{\rho^2}
    \eeta(\yy) ds_\yy.
\end{align}

@@

 Figure \ref{fig1} shows preliminary results of multiple bodies undergoing mechanical erosion in Stokes flow. The figure shows 80 bodies of initially random size and position, undergoing erosion. This figure extends previously reported 50-body simulations to 80 bodies~\cite{Quaife2018} due to a recently developed Barycentric interpolation scheme~\cite{bar2014, bar-wu-vee2015}, which enables closer contact between bodies and thus more bodies.
The color shows the (normalized) vorticity field surrounding the bodies. Vorticity, since it reduces to shear on solid interfaces, provides a convenient way to visualize local erosion rates, as well as the local flow intensity. Observe that erosion not only reduces the size of solid bodies, but also alters their shapes substantially. The bodies tend to become somewhat polygonal: corners develop connected by relatively flat faces. The number of faces does not appear to be easily predicted, but rather depends on the interaction with neighboring bodies as mediated by the Stokes flow. Notice that, as erosion proceeds, relatively straight channels tend to develop between the bodies. Certain channels that are initially larger can transmit more flow, which promotes local erosion rates and further widens these selected channels. In this way, erosion creates a runaway process in which small differences in initial channel size become amplified. The channelization process is made more evident by visualization of the velocity magnitude, as in Fig.~\ref{fig2}.

@@

An immediate goal of the proposed work is extend the range of physical effects captured in these simulations. A particular challenge of a BIE formulation is developing quadrature for nearly-singular integrands that arise when bodies are close. To address this problem, we have very recently extended a Barycentric quadrature~\cite{bar2014, bar-wu-vee2015} and have successfully simulated erosion with grains that are initially separated by 1\% of an arclength spacing. This quadrature approach has the advantage that it is a non-intrusive modification of the trapezoid rule, allowing us to achieve an error bound that independent of the grain configuration. Near-term plans are to include flow transport and gravitational sedimentation of the eroding grains. For this extension, contact forces between bodies must be included to eliminate unphysical contact that is inevitable due to discretization errors.

@@

Next, we plan to extend the simulation framework to handle chemical dissolution of solid bodies via physical laws \eqref{Ceq}--\eqref{DissVn}. For this goal, much of the computational infrastructure, such as the boundary-integral solver for the Stokes equations, is already in place, but the coupling to the advection-diffusion equations~\eqref{Ceq} and interface law~\eqref{DissVn} would need to be implemented. Once these pieces are in place, simulations of simultaneous erosion and dissolution, while also including transport and sedimentation, would finally become possible. Such a framework could accurately simulate the geophysically relevant scenario of karst conduits with embedded granular media. The gradual changes in the medium and karst boundaries due to erosion and dissolution could create conditions favorable for mechanical failure or collapse of the structure, for example sinkholes.
\todo[inline]{If we are going to propose the Laplace transform idea
to solve advection-diffusion, this might be the place to put it}

@@

\begin{wrapfigure}[]{r}{0.3\textwidth}
  \includegraphics[width=0.3\textwidth]{figs/100b_t100tracer}
  \caption{\label{fig:100tracers} A visualization of the streamlines
  with 200 tracers passing through an eroded geometry with a porosity of
  0.63. The tracers are initialized at the left end of the channel and
  the flow is driven by a pressure difference from left to right.}
\end{wrapfigure}

Transport through eroded media is a direct consequence of the pore structure. Qualitative properties of an eroded media, such as channels of high porosity, can be modest in volume fraction, but can be responsible for a large portion of the flux~\cite{Quaife2018}.  To increase understanding of flow through eroded media, we will simulate and statistically analyze transport through eroded geometries.

In an eroded geometry $\Omega$, the dimensionless
concentration $c$ is governed by
\begin{align}
  \pd{c}{t} + \uu \cdot \nabla c = \frac{1}{\Pe} \Delta c, 
  \label{eqn:advectionDiffusion}
\end{align}
where $\Pe$ is the dimensionless Peclet number. Here we assume that the
concentration $c$ is affected by the eroded geometry, but the geometry
is not altered further by the concentration. In the infinite Peclet
number limit, the concentration is governed by pure advection and the
characteristics of the concentration field are identical to the
streamlines (Figure~\ref{fig:100tracers}).

The $j^{th}$ trajectory, or characteristic, is governed by $\dot{\ss}_j(t) = \uu(\ss_j(t))$. Once these trajectories are computed, using for example a high-order Runge-Kutta method, a statistical analysis can be used to describe the transport. Two of the most important measurements are the tortuosity and the anomalous dispersion, both which are defined in terms of the length of the trajectories \begin{align}
  \lambda_j(t) = \int_{0}^{t} \|\ss'(t')\| dt'.
\end{align}
Upon computing $N_p$ trajectories, the tortuosity is the dimensionless number
\begin{align}
  T = \frac{1}{d} \left(\int_{S} u_1(x_0,y_0)\lambda(y_0) dy_0\, \right)
  \Bigg/ \left(\int_S u_1(x_0,y_0)\, dy_0 \right),
\end{align}
where $S$ is the vertical cross-section at $x=x_0$, $d$ is its length,
$u_1$ is the horizontal velocity, and $\lambda(y_0)$ is the length of
the trajectory starting at $(x_0,y_0)$ and ending at a vertical
cross-section at the other end of the channel. Relationships between the
tortuosity and porosity of a geometry have been developed by other
groups~\cite{kop-kat-tim1996, dud-koz-mat2011, mat-kha-koz2008}, and
preliminary results indicate that such models might be appropriate for
geometries whose porosity is increasing under the process of erosion
(Figure~\ref{fig:100tortuosity}). However, preliminary results also
indicate that erosion can increase the amount of tortuosity as is
evident towards the at the towards the end of the erosion process in
Figure~\ref{fig:100tortuosity}. Moreover, we have observed a significant
increase in the tortuosity at small porosities
(Figure~\ref{fig:20tortuosity}.

\begin{figure}[htp]
  \begin{center}
  \begin{subfigure}[b]{0.45\textwidth}
  \includegraphics[width=\textwidth]{figs/tort_eulerian20}
  \caption{20 eroding bodies}
  \end{subfigure}
  \begin{subfigure}[b]{0.45\textwidth}
  \includegraphics[width=\textwidth]{figs/tort_eulerian100}
  \caption{100 eroding bodies}
  \end{subfigure}
  \end{center}

  \caption{\label{fig:100tortuosity} The tortuosity of two eroding
  geometries. The tortuosity can be computed using two different formulas
  (red vs.~blue marks). The red square is the tortuosity in
  Figure~\ref{fig:100tracers}. Their is qualitative agreement between
  our numerical simulations and a power-law model (black dashed
  line)~\cite{mat-kha-koz2008}.}
  \todo[inline]{Fix caption and its references}
\end{figure}

While the tortuosity characterizes the amount of winding of streamlines, the dispersion characterizes the spreading of the trajectories.  Dispersion can be described using the first- and second-ensemble moments of the trajectories
\begin{align}
  \langle \lambda \rangle (t) = \frac{1}{N_p} \sum_{j=1}^{N_p}
    \lambda_j(t), \quad \sigma_\lambda^{2}(t) = \frac{1}{N_p}
    \sum_{j=1}^{N_p} (\lambda_j(t) - \langle \lambda \rangle(t))^2
\end{align}
Then, the standard deviation $\sigma_\lambda$ characterizes the
dispersion, and long-time behavior typically results in a power-law
scaling $\sigma_\lambda \sim t^{\alpha}$. The power $\alpha$, which
quantifies the long-time spreading, is $1$ in a ballistic regime which
occurs at early times when the trajectories have not explored much of
the porous geometry. However, once the trajectories have traversed a few
grains, dispersion in porous media is often super-dispersive with
$\alpha \in (0.5,1)$~\cite{}. In Figure~\ref{fig:100dispersion}, we
investigate spreading at several different porosities of a porous
geometry that is initialized with 100 grains.
\begin{wrapfigure}[]{r}{0.4\textwidth}
  \includegraphics[width=0.4\textwidth]{figs/100b_second_moment_long_ref}
  \caption{\label{fig:100dispersion} The particle spreading,
  characterized with the standard deviation $\sigma_\lambda(t)$ at 6
  different porosities. At early times, the line of slope 1 (black
  curve) indicates a ballistic motion. Then, at later times, the
  dispersion becomes super-dispersive at a rate indicated by the slope
  of the black dashed-dotted line.}
\end{wrapfigure}

@@

	In very preliminary work, PI Moore has been running simplified simulations with a current graduate student. These preliminary simulations are based on a hybrid continuum-discrete model for the permeability field and grain particle field respectively. They utilize the homogenized description of the flow field in \eqref{DarcyEq} and incorporate the physical effects of:
\begin{itemize}[noitemsep]
\item Transport of grains due to seepage force from the surrounding fluid flow.
\item Gravitational sedimentation of grains.
\item Feedback of grain-distribution onto permeability field.
\item Contact and cohesive forces between grains.
\end{itemize}
Preliminary investigation suggests that this last mechanisms of cohesion is essential to model sinkhole collapse in the system.
